---
title: "glp1a_epigenetics"
output: html_document
date: "2024-07-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacRatTraining6mo)
library(tidyr)
library(dplyr)
library(conflicted)
library(ggplot2)
library(tibble)
library(ComplexHeatmap)
library(here)
library(ggbreak)
library(patchwork)
library(circlize)
library(Biobase)
library(gridExtra)
library(ggpubr)

output_loc = here()
source(file.path(output_loc, "glp1a_helpers.R"))
```

# this report is for evaluating epigenetic relationships at a single feature level

So our goal is to try to transition things into the single feature level per individual rat sample, instead of commenting on things on a group level. 

This may help us more clearly see which features are actually meaningfully regulated by the epigenetics that we're describing here. 

```{r setting up data, message = F}
interested_genes = c('GCGR', 'GLP1R', 'GLP2R', 'GIPR', 'SCTR', 'VIPR1', 'VIPR2', 'NPY2R', 'CCKAR', 'CCKBR')


tissue_types_1b = MotrpacRatTraining6moData::TISSUE_ABBREV
tissue_types_1b = tissue_types_1b[tissue_types_1b != "PLASMA"]
# all_da_results_subset_filter = find_interested_features()
all_da_results_subset_filter = readRDS(file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS"))

# mapping = all_da_results_subset_filter %>% select(feature_ID, gene_symbol) %>% distinct(feature_ID, .keep_all = T)
# qc_norm_list = list()
# for(tissue in tissue_types_1b){
#   combined_qc_norm = combine_normalized_data(tissues = tissue, assays = c("ATAC", "METHYL", "TRNSCRPT"), include_epigen = T, exclude_outliers = T)
#   if(!"ATAC" %in% combined_qc_norm$assay | !"METHYL" %in% combined_qc_norm$assay) next
#   combined_qc_norm_annotated = left_join(combined_qc_norm, mapping, by = "feature_ID") %>%
#     filter(!is.na(gene_symbol)) %>%
#     select(-feature, -tissue)
#   qc_norm_list[[tissue]] = combined_qc_norm_annotated
#   gc()
# }
# saveRDS(qc_norm_list, file = paste0(repo_dir, "data/glp1a_all_omes_qc_norm_subset.RDS"))
qc_norm_list = readRDS(file.path(here(), "data", "glp1a_all_omes_qc_norm_subset.RDS"))
```


So here we squish the values by `gene_symbol`, which means we basically just
average all the epigenetic corresponding features/sites to the mean within that 
corresponding gene_symbol. The mapping for a site is determined by the 
Motrpac consortium when the data is loaded.
```{r}
mean_median_epigen = all_da_results_subset_filter %>%
  group_by(gene_symbol, assay, tissue, sex, comparison_group) %>%
  mutate(
    mean_logFC = mean(logFC, na.rm = TRUE),
    median_logFC = median(logFC, na.rm = TRUE)
  ) %>%
  select(-feature_ID, -p_value, -adj_p_value, -logFC, -zscore) %>%
  distinct(assay, tissue, sex, comparison_group, gene_symbol, .keep_all = T) %>%
  arrange(gene_symbol,tissue, sex, comparison_group, assay) %>%
  pivot_wider(
    names_from = c(assay),
    values_from = c(mean_logFC, median_logFC)
  ) %>%
  filter(!is.na(mean_logFC_TRNSCRPT) & !is.na(mean_logFC_ATAC) & !is.na(mean_logFC_METHYL))
```

Note: these plots include all of our "interested features" - so any confidence
intervals are going to built off all of these features 
```{r overall trsncript vs atac}
rna_atac = ggplot(mean_median_epigen, aes(x = mean_logFC_ATAC, y = mean_logFC_TRNSCRPT, color = tissue, shape = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Mean LogFC ATAC", y = "Mean LogFC Trnspt", title = "TRNSCRPT vs ATAC")
rna_atac
```
Note: these plots include all of our "interested features" - so any confidence
intervals are going to built off all of these features 
```{r overall trsncript vs methyl}
rna_methyl = ggplot(mean_median_epigen, aes(x = mean_logFC_METHYL, y = mean_logFC_TRNSCRPT, color = tissue, shape = sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Mean LogFC Methyl", y = "Mean LogFC Trns", title = "TRNSCRPT vs METHYL") 
rna_methyl
```

So instead of the above, we just make individual data points for sites, instead
of averaging across sites for a given gene mapping. 

Note: these plots include all of our "interested features"
```{r epigen not averaging}
# here we let multiple individual features corresponding to the same gene plot
# above we just averaged if multiple features belonged to the same gene
all_da_results_subset_sig_filter = all_da_results_subset_filter %>%
  filter(assay == "ATAC" | assay == "TRNSCRPT") %>%
  dplyr::filter(p_value < 0.05) 
atac_data = all_da_results_subset_sig_filter %>%
  filter(assay == "ATAC") %>%
  select(gene_symbol, logFC, sex, comparison_group, tissue, p_value) %>%
  rename(logFC_ATAC = logFC)
trnscrpt_data = all_da_results_subset_filter %>%
  filter(assay == "TRNSCRPT") %>%
  select(gene_symbol, logFC, sex, comparison_group, tissue, p_value) %>%
  rename(logFC_TRNSCRPT = logFC)
#i am confident there's a better way to organize this but the above works...
plot_data_atac = atac_data %>%
  inner_join(trnscrpt_data, by = c("gene_symbol", "sex", "comparison_group", "tissue")) 

scatter_plot_atac = ggplot(plot_data_atac, aes(x = logFC_ATAC, y = logFC_TRNSCRPT, shape = sex, color = comparison_group)) +
  geom_point() +
  labs(x = "LogFC in ATAC", y = "LogFC in TRNSCRPT", title = "LogFC in ATAC vs TRNSCRPT all SITES") 

print(scatter_plot_atac)
```
Note: these plots include all of our "interested features"
```{r all sites}
plot_data_atac_tissue = plot_data_atac %>%
  group_by(tissue) %>%
  mutate(correlation = cor(logFC_ATAC, logFC_TRNSCRPT, method = "pearson")) %>%
  ungroup()

# Create the scatter plot with facets by tissue and correlation label
scatter_plot_atac = ggplot(plot_data_atac_tissue, aes(x = logFC_ATAC, y = logFC_TRNSCRPT, shape = sex, color = comparison_group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") + 
  labs(x = "LogFC in ATAC", y = "LogFC in TRNSCRPT", title = "LogFC in ATAC vs TRNSCRPT, all site by tissue") +
  facet_wrap(~tissue) + 
  geom_text(aes(label = paste0("Corr overall: ", round(correlation, 2))), x = Inf, y = Inf, hjust = 1.1, vjust = 1.1)

print(scatter_plot_atac)
```
Note: these plots include all of our "interested features"
```{r all site by tissue and sex}
plot_data_atac = plot_data_atac %>%
  group_by(tissue, sex) %>%
  mutate(correlation = cor(logFC_ATAC, logFC_TRNSCRPT, method = "pearson")) %>%
  ungroup()

scatter_plot_atac = ggplot(plot_data_atac, aes(x = logFC_ATAC, y = logFC_TRNSCRPT, shape = sex, color = comparison_group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") + # optional: add trend line
  labs(x = "LogFC in ATAC", y = "LogFC in TRNSCRPT", title = "LogFC in ATAC vs TRNSCRPT, all site by tissue and sex") +
  facet_grid(tissue ~ sex) + 
  geom_text(aes(label = paste0("Corr, per sex: ", round(correlation, 2))), x = Inf, y = Inf, hjust = 1.1, vjust = 1.1)

print(scatter_plot_atac)
#note that these n start to get small if we're doing correlations per sex. its a little bit better because we're doing it site-level but still have to keep the sample size in mind. 
```

# we repeat the above plots by with methyl vs transcript now
```{r}
all_da_results_subset_sig_filter = all_da_results_subset_filter %>%
  dplyr::filter(p_value < 0.05) 
# unique(all_da_results_subset_sig_filter$tissue)

methyl_data = all_da_results_subset_sig_filter %>%
  filter(assay == "METHYL") %>%
  select(gene_symbol, logFC, sex, comparison_group, tissue, p_value) %>%
  rename(logFC_methyl = logFC)

trnscrpt_data = all_da_results_subset_filter %>%
  filter(assay == "TRNSCRPT") %>%
  select(gene_symbol, logFC, sex, comparison_group, tissue, p_value) %>%
  rename(logFC_TRNSCRPT = logFC)

# Duplicate TRNSCRPT logFC values to match each corresponding ATAC value for the same gene_symbol
plot_data_methyl = methyl_data %>%
  inner_join(trnscrpt_data, by = c("gene_symbol", "sex", "comparison_group", "tissue")) 

scatter_plot_methyl = ggplot(plot_data_methyl, aes(x = logFC_methyl, y = logFC_TRNSCRPT, shape = sex, color = comparison_group)) +
  geom_point() +
  labs(x = "LogFC in Methyl", y = "LogFC in TRNSCRPT", title = "LogFC in Methyl vs TRNSCRPT, site level") 

print(scatter_plot_methyl)
```


```{r}
plot_data_methyl = plot_data_methyl %>%
  group_by(tissue) %>%
  mutate(correlation = cor(logFC_methyl, logFC_TRNSCRPT, method = "pearson")) %>%
  ungroup()

scatter_plot_methyl = ggplot(plot_data_methyl, aes(x = logFC_methyl, y = logFC_TRNSCRPT, shape = sex, color = comparison_group)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "black") + # optional: add trend line
  labs(x = "LogFC in Methyl", y = "LogFC in TRNSCRPT", title = "LogFC in Methyl vs TRNSCRPT site-level, by tissue") +
  facet_wrap(~tissue) +
  geom_text(aes(label = paste0("Correlation: ", round(correlation, 2))), x = Inf, y = Inf, hjust = 1.1, vjust = 1.1)

print(scatter_plot_methyl)
```

Now lets plot individual samples using average of all corresponding sites

These are at a sample level!
```{r}
cor_results = data.frame(
  Tissue = character(),
  Gene = character(),
  DataType = character(),
  Correlation_R = numeric(),
  cor_p_value = numeric(),
  stringsAsFactors = FALSE
)

for(tissue in names(qc_norm_list)){
  qc_norm_matrix = qc_norm_list[[tissue]]
  for(gene in unique(qc_norm_matrix$gene_symbol)){
    subset_pheno = PHENO %>% select(pid, sacrificetime,sex)
    specific_feat = qc_norm_matrix %>% filter(gene_symbol == gene)
    rna = specific_feat %>% filter(assay == "TRNSCRPT")
    if(nrow(rna) == 0) next
    
    atac = specific_feat %>% filter(assay == "ATAC")
    col_means = atac %>% summarise(across(where(is.numeric), mean)) %>%
      mutate(assay = "ATAC_MEAN")
    atac = bind_rows(atac, col_means, rna) %>% filter(assay == "ATAC_MEAN" | assay == "TRNSCRPT")
    atac_plot = t(atac %>% select(-feature_ID, -assay, -gene_symbol)) %>% as.data.frame()
    colnames(atac_plot) = c("ATAC_MEAN", "TRNSCRPT")
    atac_plot$pid = as.numeric(rownames(atac_plot))
    atac_plot = left_join(atac_plot, subset_pheno, by = "pid") %>% distinct(pid, .keep_all = T)
    atac_plot = atac_plot[complete.cases(atac_plot),]
    # Calculate ATAC correlation
    atac_cor = cor.test(atac_plot$ATAC_MEAN, atac_plot$TRNSCRPT, method = "pearson")
    cor_results = cor_results %>%
      add_row(Tissue = tissue, Gene = gene, DataType = "ATAC", 
              Correlation_R = atac_cor$estimate, cor_p_value = atac_cor$p.value)
    
    # Repeat for methyl
    methyl = specific_feat %>% filter(assay == "METHYL")
    col_means = methyl %>% summarise(across(where(is.numeric), mean)) %>%
      mutate(assay = "METHYL_MEAN")
    methyl = bind_rows(methyl, col_means, rna) %>% filter(assay == "METHYL_MEAN" | assay == "TRNSCRPT")
    methyl_plot = t(methyl %>% select(-feature_ID, -assay, -gene_symbol)) %>% as.data.frame() 
    colnames(methyl_plot) = c("METHYL_MEAN", "TRNSCRPT")
    methyl_plot$pid = as.numeric(rownames(methyl_plot))
    methyl_plot = left_join(methyl_plot, subset_pheno, by = "pid") %>% distinct(pid, .keep_all = T)
    methyl_plot = methyl_plot[complete.cases(methyl_plot),]
    # Calculate Methyl correlation
    methyl_cor = cor.test(methyl_plot$METHYL_MEAN, methyl_plot$TRNSCRPT, method = "pearson")
    # Store Methyl correlation results
    cor_results = cor_results %>%
      add_row(Tissue = tissue, Gene = gene, DataType = "METHYL", 
              Correlation_R = methyl_cor$estimate, cor_p_value = methyl_cor$p.value)
    
    if(abs(atac_cor$estimate) < 0.24 & abs(methyl_cor$estimate) < 0.24) next 
    #if the correlations are that low, just dont bother plotting
    
    atac = ggplot(atac_plot, aes(x = ATAC_MEAN, y = TRNSCRPT)) +
      geom_point(aes(color = sacrificetime, shape = sex)) +
      geom_smooth(method = "lm", color = "darkblue", se = TRUE) + # Single confidence interval
      labs(title = paste(tissue, "ATAC:", gene),
           x = "ATAC_MEAN",
           y = "TRNSCRPT") +
      theme_minimal() +
      annotate("text", x = Inf, y = Inf, label = paste("R =", round(atac_cor$estimate, 2), 
                                                       "p =", signif(atac_cor$p.value, 2)), 
               hjust = 1.1, vjust = 1.1, size = 5, color = "black")
    ggsave(filename = file.path(here(), 
                                "figures",
                                "epigen_transcript",
                                paste0(tissue, "_", gene, "_atac.pdf")),
           width = 5, height = 5, units = "in")
    
    # Create Methyl plot with a single confidence interval and points colored by sacrificetime and shaped by sex
    methyl = ggplot(methyl_plot, aes(x = METHYL_MEAN, y = TRNSCRPT)) +
      geom_point(aes(color = sacrificetime, shape = sex)) +
      geom_smooth(method = "lm", color = "darkred", se = TRUE) + # Single confidence interval
      labs(title = paste(tissue,"Methyl:", gene),
           x = "METHYL_MEAN",
           y = "TRNSCRPT") +
      theme_minimal() +
      annotate("text", x = Inf, y = Inf, label = paste("R =", round(methyl_cor$estimate, 2), 
                                                       "p =", signif(methyl_cor$p.value, 2)), 
               hjust = 1.1, vjust = 1.1, size = 5, color = "black")
    ggsave(filename = file.path(here(), 
                                "figures",
                                "epigen_transcript",
                                paste0(tissue, "_", gene, "_methyl.pdf")),
           width = 5, height = 5, units = "in")
    
    # atac
    gridExtra::grid.arrange(atac, methyl, ncol = 2)
  }
}
```

```{r}
cor_results = cor_results %>% arrange(abs(Correlation_R))
tail(cor_results)
```
Now lets plot individual samples using average of all corresponding - now just at 8wks

Note: these are still averaged across all corresponding sites
```{r}
cor_8wks = data.frame(
  Tissue = character(),
  Gene = character(),
  DataType = character(),
  Correlation_R = numeric(),
  cor_p_value = numeric(),
  stringsAsFactors = FALSE
)

for(tissue in names(qc_norm_list)){
  qc_norm_matrix = qc_norm_list[[tissue]]
  for(gene in unique(qc_norm_matrix$gene_symbol)){
    subset_pheno = PHENO %>% select(pid, sacrificetime,sex)
    
    specific_feat = qc_norm_matrix %>% filter(gene_symbol == gene)
    rna = specific_feat %>% filter(assay == "TRNSCRPT"); if(nrow(rna) == 0) next
    
    atac = specific_feat %>% filter(assay == "ATAC")
    col_means = atac %>% summarise(across(where(is.numeric), mean)) %>%
      mutate(assay = "ATAC_MEAN")
    atac = bind_rows(atac, col_means, rna) %>% filter(assay == "ATAC_MEAN" | assay == "TRNSCRPT")
    atac_plot = t(atac %>% select(-feature_ID, -assay, -gene_symbol)) %>% as.data.frame()
    colnames(atac_plot) = c("ATAC_MEAN", "TRNSCRPT")
    atac_plot$pid = as.numeric(rownames(atac_plot))
    atac_plot = left_join(atac_plot, subset_pheno, by = "pid") %>% distinct(pid, .keep_all = T) %>%
      filter(sacrificetime == "8w")
    atac_plot = atac_plot[complete.cases(atac_plot),]
    # Calculate ATAC correlation
    atac_cor = cor.test(atac_plot$ATAC_MEAN, atac_plot$TRNSCRPT, method = "pearson")
    cor_8wks = cor_8wks %>%
      add_row(Tissue = tissue, Gene = gene, DataType = "ATAC", 
              Correlation_R = atac_cor$estimate, cor_p_value = atac_cor$p.value)
    
    # Repeat for methyl
    methyl = specific_feat %>% filter(assay == "METHYL")
    col_means = methyl %>% summarise(across(where(is.numeric), mean)) %>%
      mutate(assay = "METHYL_MEAN")
    methyl = bind_rows(methyl, col_means, rna) %>% filter(assay == "METHYL_MEAN" | assay == "TRNSCRPT")
    methyl_plot = t(methyl %>% select(-feature_ID, -assay, -gene_symbol)) %>% as.data.frame() 
    colnames(methyl_plot) = c("METHYL_MEAN", "TRNSCRPT")
    methyl_plot$pid = as.numeric(rownames(methyl_plot))
    methyl_plot = left_join(methyl_plot, subset_pheno, by = "pid") %>% distinct(pid, .keep_all = T) %>%
      filter(sacrificetime == "8w")
    methyl_plot = methyl_plot[complete.cases(methyl_plot),]
    # Calculate Methyl correlation
    methyl_cor = cor.test(methyl_plot$METHYL_MEAN, methyl_plot$TRNSCRPT, method = "pearson")
    
    # Store Methyl correlation results
    cor_8wks = cor_8wks %>%
      add_row(Tissue = tissue, Gene = gene, DataType = "METHYL", 
              Correlation_R = methyl_cor$estimate, cor_p_value = methyl_cor$p.value)
    if(abs(atac_cor$estimate) < 0.3 & abs(methyl_cor$estimate) < 0.3) next #if the correlations are that low, just dont bother plotting
    
    # Create ATAC plot with confidence interval and correlation
    atac = ggplot(atac_plot, aes(x = ATAC_MEAN, y = TRNSCRPT)) +
      geom_point(aes(shape = sex)) +
      geom_smooth(method = "lm", color = "darkblue", se = TRUE) + # Add confidence interval
      labs(title = paste(tissue, "ATAC (8wk only):", gene),
           x = "ATAC_MEAN",
           y = "TRNSCRPT") +
      theme_minimal() +
      stat_cor(method = "pearson", label.y.npc = "top")
    ggsave(filename = file.path(here(), 
                                "figures",
                                "epigen_transcript",
                                paste0(tissue, "_", gene, "_atac_wk8.pdf")),
           width = 5, height = 5, units = "in")
    
    methyl = ggplot(methyl_plot, aes(x = METHYL_MEAN, y = TRNSCRPT)) +
      geom_point(aes(shape = sex)) +
      geom_smooth(method = "lm", color = "darkred", se = TRUE) + # Add confidence interval
      labs(title = paste(tissue, "Methyl (8wk only):", gene),
           x = "METHYL_MEAN",
           y = "TRNSCRPT") +
      theme_minimal() +
      stat_cor(method = "pearson", label.y.npc = "top") 
    ggsave(filename = file.path(here(), 
                                "figures",
                                "epigen_transcript",
                                paste0(tissue, "_", gene, "_methyl_wk8.pdf")),
           width = 5, height = 5, units = "in")
    
    # Print the plots together
    gridExtra::grid.arrange(atac, methyl, ncol = 2)
  }
}
```



```{r}
all_da_results_8wks = all_da_results_subset_filter %>% filter(comparison_group == "8w")

cor_8wks_plot = cor_8wks %>% arrange(abs(Correlation_R)) %>%
  rename(tissue = Tissue) %>%
  rename(gene_symbol = Gene) %>% 
  rename(assay = DataType)
cor_8wks_plot = left_join(cor_8wks_plot, all_da_results_8wks, by = c("tissue", "gene_symbol", "assay"))
tail(cor_8wks_plot)
```

So one of the most interesting findings is CCKBR in HIPPOC -> lets go ahead
and plot the relationship at each of the timepoints
```{r}
for(single_interested_gene in c("Cckbr", "Cckar")) {
  single_interested_tissue = "HIPPOC"
  qc_norm_matrix = qc_norm_list[[single_interested_tissue]]
  
  subset_pheno = MotrpacRatTraining6moData::PHENO %>% select(pid, sacrificetime, sex)
  specific_feat = qc_norm_matrix %>% filter(gene_symbol == single_interested_gene)
  rna = specific_feat %>% filter(assay == "TRNSCRPT")
  atac = specific_feat %>% filter(assay == "ATAC")
  col_means = atac %>%
    summarise(across(where(is.numeric), mean)) %>%
    mutate(assay = "ATAC_MEAN")
  
  atac = bind_rows(atac, col_means, rna) %>%
    filter(assay %in% c("ATAC_MEAN", "TRNSCRPT"))
  
  atac_plot = t(atac %>% select(-feature_ID, -assay, -gene_symbol)) %>%
    as.data.frame()
  colnames(atac_plot) = c("ATAC_MEAN", "TRNSCRPT")
  atac_plot$pid = as.numeric(rownames(atac_plot))
  atac_plot = left_join(atac_plot, subset_pheno, by = "pid") %>%
    distinct(pid, .keep_all = TRUE)
  atac_plot = atac_plot[complete.cases(atac_plot),]
  
  plot_list = list()
  
  #-------------------------------------------
  # Combined plot (all timepoints)
  #-------------------------------------------
  combined_cor = cor.test(atac_plot$ATAC_MEAN, atac_plot$TRNSCRPT, method = "pearson")
  
  combined_plot = ggplot(atac_plot, aes(x = ATAC_MEAN, y = TRNSCRPT)) +
    geom_point(aes(color = sex, shape = sex), size = 2) +
    geom_smooth(method = "lm", color = "darkblue", se = TRUE) +
    labs(
      title = paste0("All Timepoints, N = ", nrow(atac_plot)),
      # title = paste0("All Timepoints (R =", round(combined_cor$estimate, 2),
      #                ", p =", signif(combined_cor$p.value, 2), ")"),
      x = "ATAC_MEAN",
      y = "TRNSCRPT",
      color = "Sex",
      shape = "Sex"
    ) +
    theme_minimal(base_size = 10) +
    stat_cor(method = "pearson", label.y.npc = "top") +
    guides(shape = "none") +
    theme(legend.position = "bottom") +
    scale_color_manual(values = c("male" = "#1F77B4",
                                  "female" = "#E377C2"))
  
  plot_list[["All"]] = combined_plot
  
  #-------------------------------------------
  # Individual timepoint plots
  #-------------------------------------------
  for(tp in c("8w")) { #"1w", "2w", "4w",
    atac_plot_tp = atac_plot %>% filter(sacrificetime == tp)
    atac_cor = cor.test(atac_plot_tp$ATAC_MEAN, atac_plot_tp$TRNSCRPT, method = "pearson")
    
    atac_tp = ggplot(atac_plot_tp, aes(x = ATAC_MEAN, y = TRNSCRPT)) +
      geom_point(aes(color = sex, shape = sex), size = 2) +
      geom_smooth(method = "lm", color = "darkblue", se = TRUE) +
      labs(
        title = paste0(tp, " Only, N = ", nrow(atac_plot_tp)),
        # title = paste0(tp, " Only (R =", round(atac_cor$estimate, 2),
        #               ", p =", signif(atac_cor$p.value, 2), ")"),
        x = "ATAC_MEAN",
        y = "TRNSCRPT",
        color = "Sex",
        shape = "Sex"
      ) +
      theme_minimal(base_size = 10) +
      stat_cor(method = "pearson", label.y.npc = "top") +
      guides(shape = "none") +
      theme(legend.position = "bottom") +
      scale_color_manual(values = c("male" = "#1F77B4",
                                    "female" = "#E377C2"))
    
    plot_list[[tp]] = atac_tp
  }
  
  #-------------------------------------------
  # Save combined figure (5 columns) -> new plan just include 5 as a sup. 
  # to toggle b/t all tp and just 2 see above for tp in c(...), switch
  # ncol = 5, width = 17
  #-------------------------------------------
  fig_name = paste0("fig_2d_", single_interested_gene, "_", single_interested_tissue, ".pdf")
  pdf(file = file.path(here(), "figures", fig_name),
      width = 6,  # wider for 5 columns
      height = 4)
  
  gridExtra::grid.arrange(
    grobs = plot_list,
    ncol = 2,
    top = paste0(single_interested_tissue, " — ", single_interested_gene),
    respect = TRUE
  )

  dev.off()
}

```

CCKBR ATAC MEAN vs TRNSCRPT for all timepoints and 8w only, concatenated with
patchwork so the two panels share one legend.
```{r cckbr atac mean shared legend}
single_interested_gene = "Cckbr"
single_interested_tissue = "HIPPOC"
qc_norm_matrix = qc_norm_list[[single_interested_tissue]]

subset_pheno = MotrpacRatTraining6moData::PHENO %>% select(pid, sacrificetime, sex)
specific_feat = qc_norm_matrix %>% filter(gene_symbol == single_interested_gene)
rna = specific_feat %>% filter(assay == "TRNSCRPT")
atac = specific_feat %>% filter(assay == "ATAC")
col_means = atac %>%
  summarise(across(where(is.numeric), mean)) %>%
  mutate(assay = "ATAC_MEAN")

atac = bind_rows(atac, col_means, rna) %>%
  filter(assay %in% c("ATAC_MEAN", "TRNSCRPT"))

atac_plot = t(atac %>% select(-feature_ID, -assay, -gene_symbol)) %>%
  as.data.frame()
colnames(atac_plot) = c("ATAC_MEAN", "TRNSCRPT")
atac_plot$pid = as.numeric(rownames(atac_plot))
atac_plot = left_join(atac_plot, subset_pheno, by = "pid") %>%
  distinct(pid, .keep_all = TRUE)
atac_plot = atac_plot[complete.cases(atac_plot),]

all_tp_plot = ggplot(atac_plot, aes(x = ATAC_MEAN, y = TRNSCRPT)) +
  geom_point(aes(color = sex, shape = sex), size = 2) +
  geom_smooth(method = "lm", color = "darkblue", se = TRUE) +
  labs(
    title = paste0("All Timepoints, N = ", nrow(atac_plot)),
    x = "ATAC logCPM",
    y = "Transcript logCPM",
    color = "Sex",
    shape = "Sex"
  ) +
  theme_minimal(base_size = 10) +
  stat_cor(method = "pearson", label.y.npc = "top") +
  guides(shape = "none") +
  scale_color_manual(values = c("male" = "#1F77B4",
                                "female" = "#E377C2"))

atac_plot_8w = atac_plot %>% filter(sacrificetime == "8w")
wk8_plot = ggplot(atac_plot_8w, aes(x = ATAC_MEAN, y = TRNSCRPT)) +
  geom_point(aes(color = sex, shape = sex), size = 2) +
  geom_smooth(method = "lm", color = "darkblue", se = TRUE) +
  labs(
    title = paste0("8w Only, N = ", nrow(atac_plot_8w)),
    x = "ATAC logCPM",
    y = "Transcript logCPM",
    color = "Sex",
    shape = "Sex"
  ) +
  theme_minimal(base_size = 10) +
  stat_cor(method = "pearson", label.y.npc = "top") +
  guides(shape = "none") +
  scale_color_manual(values = c("male" = "#1F77B4",
                                "female" = "#E377C2"))

cckbr_combined = (all_tp_plot + wk8_plot) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = paste0(single_interested_tissue, " — ", single_interested_gene),
    theme = theme(plot.title = element_text(hjust = 0.5))
  ) &
  theme(legend.position = "bottom")

fig_name = paste0("fig_2d_", single_interested_gene, "_", single_interested_tissue, "_shared_legend.pdf")
ggsave(filename = file.path(here(), "figures", fig_name),
       plot = cckbr_combined,
       width = 6,
       height = 4,
       units = "in")

cckbr_combined
```





