---
title: "glp1a_exploration"
author: "Christopher Jin"
date: '2023-10-24'
output: html_document
---

# General Targets and Initial Look

Incretin Target List

GCG peptide products: GCG, GLP1(ZGLP), GLP2(not available in our data set); note there are other products that form and have differing affinities for the below receptors
Receptors: GCGR, GLP1R, GLP2R
Enzymes that facilitate cleavage (PCSK2, PCSK1)

GIP
Receptor: GIPR

Other incretins of the GCG super family and their receptors:
SCT/SCTR
VIP/VIPR1/VIPR2
ADCYAP1/ADCYAP1R1

Incretin-like or regulatory factors
PYY, NPY; receptor (NPY2R)
CCK/CCKAR/CCKBR
Dpp4 (inhibitor)
SST (somatostatin degrades incretins)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacRatTraining6mo)
library(tidyr)
library(dplyr)
library(clusterProfiler)
library(KEGGREST)
library(conflicted)
library(ggplot2)
library(tibble)
library(ComplexHeatmap)
library(here)
library(ggbreak)
library(patchwork)
library(circlize)
library(Biobase)
library(gridExtra)

output_loc = here()
source(file.path(output_loc, "glp1a_helpers.R"))
```

```{r just establishing info, message = F, include = F}
interested_genes = c('GCGR', 'GLP1R', 'GLP2R', 'GIPR', 'SCTR', 'VIPR1', 'VIPR2', 'NPY2R', 'CCKAR', 'CCKBR')

select_tissue_types = MotrpacRatTraining6moData::TISSUE_ABBREV
select_tissue_types = select_tissue_types[!select_tissue_types %in% c("PLASMA", "OVARY", "TESTES")]

all_feat_to_gene = MotrpacRatTraining6moData::FEATURE_TO_GENE
trns_feat_to_gene = MotrpacRatTraining6mo::load_feature_annotation(assay = "TRNSCRPT") #doing this to just check how many in RNA specifically are represented
genes_in = trns_feat_to_gene %>%
  dplyr::filter(tolower(gene_name) %in% tolower(interested_genes))

paste(unique(genes_in$gene_name),sep = " ") # 9 genes featured

all_feat_in = all_feat_to_gene %>% #all omes corresponding to those previous features 
  dplyr::filter(ensembl_gene %in% genes_in$gene_id)

#see subseting process. splitting off here bc the loading of epigen is slow. 
# all_da_results_subset_filter = find_interested_features(select_tissue_types)
# saveRDS(all_da_results_subset_filter, file = file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS"))
all_da_results_subset_filter = readRDS(file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS")) 
```

```{r, eval = TRUE}
all_immuno_da_results = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types,
                                                                  assays = "IMMUNO",
                                                                  include_epigen = F) %>%
  mutate(gene_symbol = "") %>%
  select(feature_ID, gene_symbol, assay, tissue, sex, comparison_group, p_value, logFC) %>%
  filter(p_value < 0.05,
         feature_ID %in% c("GIP", "GLP1", "GLUCAGON"))

significant_gene_results = all_da_results_subset_filter %>%
  filter(p_value < 0.05) %>%
  select(feature_ID, gene_symbol, assay, tissue, sex, comparison_group, p_value, logFC) %>%
  arrange(assay)

full_significant_results = rbind(all_immuno_da_results, significant_gene_results)
# View(full_significant_results %>% filter(assay %in% c("TRNSCRPT", "IMMUNO")))
write.csv(full_significant_results, 
          file = file.path(here(), "data", "full_da_results.csv"),
          row.names = FALSE)
```

Protein check. None of the proteins have been sufficiently reliable detected for any of the incretin receptors, so we just avoid including proteomics in further analysis.
```{r}
all_da_results_subset = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types, assays = c("PROT")) %>%
  left_join(FEATURE_TO_GENE, by = "feature_ID") %>%
  filter(gene_symbol %in% interested_genes)
```

some of these plots end up in figure 2D
```{r visualizing norm data, eval = F, warning = F, message = F}
for(feat in unique(all_da_results_subset_filter$feature_ID)){
  for (tissue in tissue_types_1b){
    if(tissue == "OVARY" | tissue == "TESTES") next
    #use default 1B function
    desired_plot = MotrpacRatTraining6mo::plot_feature_normalized_data(assay = 'TRNSCRPT',
                                                                       tissue = tissue,
                                                                       feature_ID = feat,
                                                                       add_gene_symbol = T)
    print(desired_plot)
  }
}
```

```{r figure 1}
rna_tissues = select_tissue_types
immuno_tissues = MotrpacRatTraining6moData::TISSUE_ABBREV[!MotrpacRatTraining6moData::TISSUE_ABBREV %in% c("OVARY", "TESTES")]

# Glp1r & GLP1
rna_glp1r = get_expression_stats("ENSRNOG00000001152", rna_tissues, "TRNSCRPT")

#we're just tossing plasma into blood to match up levels with RNA
immuno_glp1 = get_expression_stats("GLP1", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

#pausing troubleshooting:: this doesn't work right now and I'm not sure why.
#it seems to make the breaks if i just return the rna plot, but if they get combined
#it doesn't work. maybe patchwork got updated or something?
glp1_fig1 = plot_rna_immuno_combined(
  rna_data = rna_glp1r,
  immuno_data = immuno_glp1,
  include_first_legend = FALSE,
  gene_name = "GLP1R",
  rna_y_breaks = list(c(2.5, 7), c(12, 75)),
  immuno_reverse = FALSE
)
pdf(file = file.path(here(), "figures", "fig_1_glp1.pdf"))
print(glp1_fig1)
dev.off()

# Gipr & GIP
rna_gipr = get_expression_stats("ENSRNOG00000015860", rna_tissues, "TRNSCRPT")
#we're just tossing plasma into blood to match up levels with RNA
immuno_gip = get_expression_stats("GIP", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

gip_fig1 = plot_rna_immuno_combined(
  rna_data = rna_gipr,
  immuno_data = immuno_gip,
  gene_name = "GIPR",
  immuno_reverse = FALSE
)

pdf(file = file.path(here(), "figures", "fig_1_gip.pdf"))
print(gip_fig1)
dev.off()
```


```{r figure 2a, eval = T, warning = F}
scale = 2
#so in general, making one big plot with every tissue becomes v difficult to read. 
#----strategy one: possibly just group vertically for like tissues
# ht_cortex_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("CORTEX", "HIPPOC", "HYPOTH"))))
# ht_fat_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("BAT", "WAT-SC"))))
# ht_gut_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("COLON", "SMLINT"))))
# ht_kidney_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("KIDNEY", "ADRNL", "LIVER"))))
# ht_muscle_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("HEART", "SKM-GN", "SKM-VL"))))

# pdf("figures/all_tissues_heatmaps_combined_gridExtra.pdf", width = 8, height = 12)
# grid.arrange(ht_cortex_grob, ht_fat_grob, ht_gut_grob, ht_kidney_grob, ht_muscle_grob, ncol = 1)
# dev.off()
#----------------------

#----strategy two: make one trimmed one - I think i like this one better?
trimmed_selection = c("ADRNL", "BAT", 
                      "BLOOD", "COLON", 
                      "CORTEX", "HEART")
heatmap_trimmed = make_a_plot_for_tissue_heatmaps(trimmed_selection)
pdf(file = file.path(here(), "figures", "fig_2a_attempts", "trimmed_combined_rna_zscore_one.pdf"),
    width = 18 * scale, height = 6 * scale, family = "ArialMT")
draw(heatmap_trimmed, merge_legends = F,
     align_heatmap_legend = "heatmap_top",
     align_annotation_legend = "heatmap_top")
dev.off()
#------------------------

#----strategy two part b: 
#so we actually end up talking about quite a few tissues in the text so im making a 2 layer one
#and we just move the screenshots together?
#i can bind them together using patchwork or something too..?
other_trimmed_selection = c("HIPPOC", "HYPOTH", "LUNG",
                            "SMLINT", "SPLEEN", "WAT-SC")
other_heatmap_trimmed = make_a_plot_for_tissue_heatmaps(other_trimmed_selection)
pdf(file = file.path(here(), "figures", "fig_2a_attempts", "trimmed_combined_rna_zscore_two.pdf"),
    width = 18 * scale, height = 6 * scale, family = "ArialMT")
draw(other_heatmap_trimmed, merge_legends = F,
     align_heatmap_legend = "heatmap_top",
     align_annotation_legend = "heatmap_top")
dev.off()
#------------------------

#------------------------just for supplement------------------------
heatmap_full = make_a_plot_for_tissue_heatmaps(c(trimmed_selection, other_trimmed_selection))
pdf(file = file.path(here(), "figures", "fig_2a_attempts", "all_tissue_combined_rna_zscore.pdf"),
    width = 18 * scale, height = 6 * scale, family = "ArialMT")
draw(heatmap_full, merge_legends = F,
     align_heatmap_legend = "heatmap_top",
     align_annotation_legend = "heatmap_top")
dev.off()
#------------------------
```

```{r figure 2b, eval = T}
cutoff = 0.05
#----here we can just find a big table of DA results if you'd like to view the tables.----
rna_selected_da_results = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types, assays = "TRNSCRPT") %>%
  filter(feature_ID %in% all_da_results_subset_filter$feature_ID) %>%
  filter(comparison_group == "8w") %>%
  select(assay, tissue, feature_ID, sex, comparison_group, p_value, logFC) %>%
  pivot_wider(names_from = sex, values_from = c(p_value, logFC), names_glue = "{sex}_{.value}") %>%
  mutate(shape = ifelse(male_logFC > 0 & female_logFC > 0, "up", "down")) %>%
  mutate(group = case_when(
    male_p_value < cutoff & (is.na(female_p_value) | female_p_value >= cutoff) ~ "male-specific",
    female_p_value < cutoff & (is.na(male_p_value) | male_p_value >= cutoff) ~ "female-specific",
    female_p_value < cutoff & male_p_value < cutoff  ~ "sex-consistent",
    TRUE ~ "neither-significant"
  ))  %>% 
  pivot_longer(cols = starts_with("male_") | starts_with("female_"), 
               names_to = c("sex", ".value"), 
               names_pattern = "(male|female)_(.*)")
#
rna_selected_da_results = merge(rna_selected_da_results, FEATURE_TO_GENE) %>%
  mutate(group = factor(group,
                        levels = c("sex-consistent", "female-specific", "male-specific", "neither-significant")))

#Because of the shapes the order cant really be forced to be in the same order as the other heatmaps?. 
#Going to stop troubleshooting because it is taking way too long.

rna_selected_da_results$gene_symbol = factor(rna_selected_da_results$gene_symbol,
                                             levels = sort(unique(rna_selected_da_results$gene_symbol)))


triangle_plots = ggplot(rna_selected_da_results, 
                        aes(y=gene_symbol, x=tissue, size=group, shape=shape)) +
  geom_point(aes(fill=group, colour=group)) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(),
        plot.background = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(linewidth=1),
        axis.text.y = element_text(colour="black"),
        axis.text.x = element_text(colour="black",
                                   angle=90, hjust=1, vjust=0.5),
        axis.title = element_blank(),
        strip.text.y.left = element_text(size = 10),
        strip.text.x = element_text(angle=90, hjust=0),
        axis.text.y.right = element_text(hjust=0),
        legend.position = "top",
        legend.box = "vertical",
        legend.justification = "right",
        legend.title = element_blank(),
        legend.margin = margin(t = 0, b = -0.3, unit='cm'),
        legend.spacing.x = unit(0.1, 'cm')) +
  scale_fill_manual(values=c("sex-consistent" = "white",
                             "neither-significant" = "brown",
                             "male-specific" = MotrpacBicQC::sex_cols[['male']],
                             "female-specific" = MotrpacBicQC::sex_cols[['female']]),
                    name = "Node type", guide="none") +
  scale_colour_manual(values=c("sex-consistent" = "black",
                               # "neither-significant" = "brown",
                               "male-specific" = MotrpacBicQC::sex_cols[['male']],
                               "female-specific" = MotrpacBicQC::sex_cols[['female']]),
                      name = "Node type", guide="none") +
  scale_shape_manual(values=c('up'=24, 'down'=25), 
                     labels=c(up="Up-regulated", down="Down-regulated"), 
                     name="Node direction") +
  scale_size_manual(values=c("sex-consistent" = 4,
                             # "neither-significant" = 1,
                             "male-specific" = 1.5,
                             "female-specific" = 3),
                    breaks = c("sex-consistent","female-specific","male-specific"),
                    name="Node type") +
  guides(shape=guide_legend(override.aes = list(size=3, fill="black")),
         size=guide_legend(override.aes = list(shape=24, 
                                               size=c(3,2.5,1.5), 
                                               fill=c("white",MotrpacBicQC::sex_cols[['female']],MotrpacBicQC::sex_cols[['male']]),
                                               colour=c("black","white","white")))) + 
  ggtitle("Responses after 8 weeks of training")


pdf(file = paste0(here(), "/figures/fig_2b_rna_triangle_direction.pdf"),
    width = 6, height = 4)
print(triangle_plots)
dev.off()
```


Making the immuno heatmaps but instead we're just doing timepoint by tissue
```{r figure 2c, message = F}
immuno_tissue_types = c(select_tissue_types, "PLASMA")
gip_heatmap = make_immuno_heatmap("GIP")$plot
glp_heatmap = make_immuno_heatmap("GLP1")$plot

combined_heatmap = gip_heatmap + glp_heatmap

pdf(file = paste0(here(), "/figures/fig_2c_gip_glp_immunoassay.pdf"),
    width = 6, height = 4)
draw(combined_heatmap, merge_legends = TRUE)
dev.off()
```






