---
title: "glp1a_exploration"
author: "Christopher Jin"
date: '2023-10-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacRatTraining6mo)
library(tidyr)
library(dplyr)
library(clusterProfiler)
library(KEGGREST)
library(conflicted)
library(ggplot2)
library(tibble)
library(ComplexHeatmap)
library(here)
library(ggbreak)
library(patchwork)
library(circlize)
library(Biobase)
library(gridExtra)

output_loc = here()
source(file.path(output_loc, "glp1a_helpers.R"))
```


```{r just establishing info, message = F, include = F}
# ------------------------------------------------------------------------------
# Data source disclaimer:
# All data used in this analysis are derived from the publicly available
# MotrpacRatTraining6moData R package (version 2.0.0), which provides curated
# multi-omic results from the MoTrPAC 6-month rat endurance and resistance
# training study.
# ------------------------------------------------------------------------------

interested_genes = c('GCGR', 'GLP1R', 'GLP2R', 'GIPR', 'SCTR', 'VIPR1', 'VIPR2', 'NPY2R', 'CCKAR', 'CCKBR')

select_tissue_types = MotrpacRatTraining6moData::TISSUE_ABBREV
select_tissue_types = select_tissue_types[!select_tissue_types %in% c("PLASMA", "OVARY", "TESTES")]

all_feat_to_gene = MotrpacRatTraining6moData::FEATURE_TO_GENE
trns_feat_to_gene = MotrpacRatTraining6mo::load_feature_annotation(assay = "TRNSCRPT") #doing this to just check how many in RNA specifically are represented
genes_in = trns_feat_to_gene %>%
  dplyr::filter(tolower(gene_name) %in% tolower(interested_genes))

paste(unique(genes_in$gene_name),sep = " ") # 9 genes featured

all_feat_in = all_feat_to_gene %>% #all omes corresponding to those previous features 
  dplyr::filter(ensembl_gene %in% genes_in$gene_id)

#see subseting process. splitting off here bc the loading of epigen is slow. 
# all_da_results_subset_filter = find_interested_features(select_tissue_types)
# saveRDS(all_da_results_subset_filter, file = file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS"))
# all_da_results_subset_filter = readRDS(file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS")) 

all_da_results_subset_filter = readRDS(file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS")) %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH"), .by = c("tissue", "assay", "comparison_group"))
```

Loading in data from the luminex assays and combining. The luminex assays had a much restricted number of features compared to the RNAseq, so I only pull GIP, GLP1, Glucagon.
```{r, eval = TRUE}
select_tissue_types_immuno = c(select_tissue_types, "PLASMA")
all_immuno_da_results = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types_immuno,
                                                                  assays = "IMMUNO",
                                                                  include_epigen = F) %>%
  filter(feature_ID %in% c("GIP", "GLP1", "GLUCAGON")) %>%
  mutate(gene_symbol = "") %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH"), .by = c("tissue", "assay", "comparison_group")) %>%
  select(feature_ID, gene_symbol, assay, tissue, sex, comparison_group, p_value, adj_p_value, logFC) %>%
  filter(adj_p_value < 0.1)

significant_gene_results = all_da_results_subset_filter %>%
  filter(adj_p_value < 0.1) %>%
  select(feature_ID, gene_symbol, assay, tissue, sex, comparison_group, p_value, adj_p_value, logFC) %>%
  arrange(assay)

full_significant_results = rbind(all_immuno_da_results, significant_gene_results)
# View(full_significant_results %>% filter(assay %in% c("TRNSCRPT", "IMMUNO")))
write.csv(full_significant_results, 
          file = file.path(here(), "data", "full_da_results.csv"),
          row.names = FALSE)
```

Protein check. None of the proteins have been sufficiently reliable detected for any of the incretin receptors, so we just avoid including proteomics in further analysis.
```{r}
all_da_results_subset = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types, assays = c("PROT")) %>%
  left_join(FEATURE_TO_GENE, by = "feature_ID") %>%
  filter(gene_symbol %in% interested_genes)
```

None of these plots are actually used in the manuscript. A few trajectory plots of normalized transcript expression. Helpful for general data visualizations, if desired
```{r visualizing norm data, eval = F, warning = F, message = F}
for(feat in unique(all_da_results_subset_filter$feature_ID)){
  for (tissue in tissue_types_1b){
    if(tissue == "OVARY" | tissue == "TESTES") next
    #use default 1B function
    desired_plot = MotrpacRatTraining6mo::plot_feature_normalized_data(assay = 'TRNSCRPT',
                                                                       tissue = tissue,
                                                                       feature_ID = feat,
                                                                       add_gene_symbol = T)
    print(desired_plot)
  }
}
```

Generating figure 1. The RNA_y_breaks ggplot syntax was giving me trouble because of the fixed/free scales, so I went with a zoom instead. 
```{r figure 1}
rna_tissues = select_tissue_types
immuno_tissues = MotrpacRatTraining6moData::TISSUE_ABBREV[!MotrpacRatTraining6moData::TISSUE_ABBREV %in% c("OVARY", "TESTES")]

# Glp1r & GLP1
rna_glp1r = get_expression_stats("ENSRNOG00000001152", rna_tissues, "TRNSCRPT")

#we're just tossing plasma into blood to match up levels with RNA
immuno_glp1 = get_expression_stats("GLP1", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

glp1_fig1 = plot_rna_immuno_combined(
  rna_data = rna_glp1r,
  immuno_data = immuno_glp1,
  include_first_legend = FALSE,
  gene_name = "GLP1R",
  rna_y_breaks = list(c(2.5, 7), c(12, 75)),
  immuno_reverse = FALSE
)
pdf(file = file.path(here(), "figures", "fig_1_glp1test.pdf"))
print(glp1_fig1)
dev.off()

# Gipr & GIP
rna_gipr = get_expression_stats("ENSRNOG00000015860", rna_tissues, "TRNSCRPT")
#we're just tossing plasma into blood to match up levels with RNA
immuno_gip = get_expression_stats("GIP", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

gip_fig1 = plot_rna_immuno_combined(
  rna_data = rna_gipr,
  immuno_data = immuno_gip,
  gene_name = "GIPR",
  immuno_reverse = FALSE
)

pdf(file = file.path(here(), "figures", "fig_1_giptest.pdf"))
print(gip_fig1)
dev.off()
```


```{r figure 1 split by sex}
rna_tissues = select_tissue_types
immuno_tissues = MotrpacRatTraining6moData::TISSUE_ABBREV[!MotrpacRatTraining6moData::TISSUE_ABBREV %in% c("OVARY", "TESTES")]

# Glp1r & GLP1
rna_glp1r = get_expression_stats_by_sex("ENSRNOG00000001152", rna_tissues, "TRNSCRPT")

#we're just tossing plasma into blood to match up levels with RNA
immuno_glp1 = get_expression_stats_by_sex("GLP1", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

glp1_fig1 = plot_rna_immuno_jitterbysex(
  rna_data = rna_glp1r,
  immuno_data = immuno_glp1,
  include_first_legend = TRUE,
  gene_name = "GLP1R",
  rna_y_breaks = c(c(8,70)),
  immuno_reverse = FALSE
)

pdf(file = file.path(here(), "figures", "fig_1_glp1_by_sex.pdf"))
print(glp1_fig1)
dev.off()

# Gipr & GIP
rna_gipr = get_expression_stats_by_sex("ENSRNOG00000015860", rna_tissues, "TRNSCRPT")
#we're just tossing plasma into blood to match up levels with RNA
immuno_gip = get_expression_stats_by_sex("GIP", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

gip_fig1 = plot_rna_immuno_jitterbysex(
  rna_data = rna_gipr,
  immuno_data = immuno_gip,
  gene_name = "GIPR",
  immuno_reverse = FALSE
)

pdf(file = file.path(here(), "figures", "fig_1_gip_by_sex.pdf"))
print(gip_fig1)
dev.off()
```


```{r figure 2B, eval = T, warning = F}
scale = 2
#so in general, making one big plot with every tissue becomes v difficult to read. 
#----strategy one: possibly just group vertically for like tissues
# ht_cortex_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("CORTEX", "HIPPOC", "HYPOTH"))))
# ht_fat_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("BAT", "WAT-SC"))))
# ht_gut_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("COLON", "SMLINT"))))
# ht_kidney_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("KIDNEY", "ADRNL", "LIVER"))))
# ht_muscle_grob = grid.grabExpr(draw(make_a_plot_for_tissue_heatmaps(c("HEART", "SKM-GN", "SKM-VL"))))

# pdf("figures/all_tissues_heatmaps_combined_gridExtra.pdf", width = 8, height = 12)
# grid.arrange(ht_cortex_grob, ht_fat_grob, ht_gut_grob, ht_kidney_grob, ht_muscle_grob, ncol = 1)
# dev.off()
#----------------------

#----strategy two: make one trimmed one - I think i like this one better?
top_selection = c("CORTEX", 
                  "HYPOTH",
                  "HIPPOC",
                  "BAT",
                  "WAT-SC",
                  "SKM-GN",
                  "SKM-VL"
)
bottom_selection = c("BLOOD", 
                     "HEART",
                     "LUNG",
                     "SPLEEN",
                     "ADRNL", 
                     "COLON", 
                     "SMLINT")

heatmap_trimmed = make_a_plot_for_tissue_heatmaps(top_selection,
                                                  interested_genes)
other_heatmap_trimmed = make_a_plot_for_tissue_heatmaps(bottom_selection,
                                                        interested_genes)
heatmap_full = heatmap_trimmed %v% other_heatmap_trimmed #concat vertically

pdf(file = file.path(here(), "figures", "fig_2a_attempts", "all_tissue_combined_rna_zscore.pdf"),
    width = 18 * scale, height = 6 * scale, family = "ArialMT")
draw(heatmap_full, merge_legends = FALSE,
     align_heatmap_legend = "heatmap_top",
     align_annotation_legend = "heatmap_top",
     ht_gap = unit(12, "mm"))
dev.off()
```

```{r figure 2C triangle plots, eval = T}
cutoff = 0.10
#----here we can just find a big table of DA results if you'd like to view the tables.----
rna_selected_da_results = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types, assays = "TRNSCRPT") %>%
  filter(feature_ID %in% all_da_results_subset_filter$feature_ID) %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH"), .by = c("tissue", "assay", "comparison_group")) %>%
  filter(comparison_group == "8w") %>%
  select(assay, tissue, feature_ID, sex, comparison_group, adj_p_value, logFC) %>%
  pivot_wider(names_from = sex, values_from = c(adj_p_value, logFC), names_glue = "{sex}_{.value}") %>%
  mutate(shape = ifelse(male_logFC > 0 & female_logFC > 0, "up", "down")) %>%
  mutate(group = case_when(
    male_adj_p_value < cutoff & (is.na(female_adj_p_value) | female_adj_p_value >= cutoff) ~ "male-specific",
    female_adj_p_value < cutoff & (is.na(male_adj_p_value) | male_adj_p_value >= cutoff) ~ "female-specific",
    female_adj_p_value < cutoff & male_adj_p_value < cutoff  ~ "sex-consistent",
    TRUE ~ "neither-significant"
  ))  %>% 
  pivot_longer(cols = starts_with("male_") | starts_with("female_"), 
               names_to = c("sex", ".value"), 
               names_pattern = "(male|female)_(.*)")
#
rna_selected_da_results = merge(rna_selected_da_results, FEATURE_TO_GENE) %>%
  mutate(group = factor(group,
                        levels = c("sex-consistent", "female-specific", "male-specific", "neither-significant")))

#Because of the shapes the order cant really be forced to be in the same order as the other heatmaps?. 
#Going to stop troubleshooting because it is taking way too long.
rna_selected_da_results$gene_symbol = factor(rna_selected_da_results$gene_symbol,
                                             levels = sort(unique(rna_selected_da_results$gene_symbol)))


triangle_plots = ggplot(
  rna_selected_da_results,
  aes(y = gene_symbol, x = tissue, size = group, shape = shape)
) +
  geom_point(aes(fill = group, colour = group)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(),
    plot.background = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(linewidth = 1),
    axis.text.y = element_text(colour = "black"),
    axis.text.x = element_text(
      colour = "black",
      angle = 90, hjust = 1, vjust = 0.5
    ),
    axis.title = element_blank(),
    strip.text.y.left = element_text(size = 10),
    strip.text.x = element_text(angle = 90, hjust = 0),
    axis.text.y.right = element_text(hjust = 0),
    legend.position = "top",
    legend.box = "vertical",
    legend.justification = "right",
    legend.title = element_blank(),
    legend.margin = margin(t = 0, b = -0.3, unit = "cm"),
    legend.spacing.x = unit(0.1, "cm")
  ) +
  scale_fill_manual(
    values = c(
      "sex-consistent" = "white",
      "male-specific" = MotrpacBicQC::sex_cols[["male"]],
      "female-specific" = MotrpacBicQC::sex_cols[["female"]]
    ),
    breaks = c("sex-consistent", "female-specific", "male-specific"),
    name = "Node type"
  ) +
  scale_colour_manual(
    values = c(
      "sex-consistent" = "black",
      "male-specific" = MotrpacBicQC::sex_cols[["male"]],
      "female-specific" = MotrpacBicQC::sex_cols[["female"]]
    ),
    breaks = c("sex-consistent", "female-specific", "male-specific"),
    name = "Node type"
  ) +
  scale_shape_manual(
    values = c(up = 24, down = 25),
    labels = c(up = "Up-regulated", down = "Down-regulated"),
    name = "Node direction"
  ) +
  scale_size_manual(
    values = c(
      "sex-consistent" = 4,
      "female-specific" = 3,
      "male-specific" = 2.5
    ),
    breaks = c("sex-consistent", "female-specific", "male-specific"),
    name = "Node type"
  ) +
  guides(
    shape = guide_legend(
      override.aes = list(size = 3, fill = "black")
    ),
    size = guide_legend(
      override.aes = list(
        shape = 24,
        size = c(3, 2.5, 1.5),
        fill = c(
          "white",
          MotrpacBicQC::sex_cols[["female"]],
          MotrpacBicQC::sex_cols[["male"]]
        ),
        colour = c("black", "white", "white")
      )
    )
  ) +
  ggtitle("Responses after 8 weeks of training")


pdf(file = paste0(here(), "/figures/fig_2c_rna_triangle_direction.pdf"),
    width = 6, height = 4)
print(triangle_plots)
dev.off()
```


Making the immuno heatmaps but instead we're just doing timepoint by tissue
```{r figure 2D Immuno heatmaps, message = F}
immuno_tissue_types = c(select_tissue_types, "PLASMA")
gip_heatmap = make_immuno_heatmap("GIP")$plot
glp_heatmap = make_immuno_heatmap("GLP1")$plot

combined_heatmap = gip_heatmap + glp_heatmap

pdf(file = paste0(here(), "/figures/fig_2D_gip_glp_immunoassay.pdf"),
    width = 6, height = 4)
draw(combined_heatmap, merge_legends = TRUE)
dev.off()
```






