---
title: "glp1a_exploration"
author: "Christopher Jin"
date: '2023-10-24'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MotrpacRatTraining6mo)
library(tidyr)
library(dplyr)
library(clusterProfiler)
library(KEGGREST)
library(conflicted)
library(ggplot2)
library(tibble)
library(ComplexHeatmap)
library(here)
library(ggbreak)
library(patchwork)
library(circlize)
library(Biobase)
library(gridExtra)
library(gtexr)
library(kableExtra)

output_loc = here()
source(file.path(output_loc, "glp1a_helpers.R"))
```


```{r just establishing info, message = F, include = F}
# ------------------------------------------------------------------------------
# Data source disclaimer:
# All data used in this analysis are derived from the publicly available
# MotrpacRatTraining6moData R package (version 2.0.0), which provides curated
# multi-omic results from the MoTrPAC 6-month rat endurance and resistance
# training study.
# ------------------------------------------------------------------------------

interested_genes = c('GCGR', 'GLP1R', 'GLP2R', 'GIPR', 'SCTR', 'VIPR2', 'NPY2R', 'CCKAR', 'CCKBR')

select_tissue_types = MotrpacRatTraining6moData::TISSUE_ABBREV
select_tissue_types = select_tissue_types[!select_tissue_types %in% c("PLASMA", "OVARY", "TESTES")]

all_feat_to_gene = MotrpacRatTraining6moData::FEATURE_TO_GENE
trns_feat_to_gene = MotrpacRatTraining6mo::load_feature_annotation(assay = "TRNSCRPT") #doing this to just check how many in RNA specifically are represented
genes_in = trns_feat_to_gene %>%
  dplyr::filter(tolower(gene_name) %in% tolower(interested_genes))

paste(unique(genes_in$gene_name),sep = " ") # 9 genes featured

all_feat_in = all_feat_to_gene %>% #all omes corresponding to those previous features 
  dplyr::filter(ensembl_gene %in% genes_in$gene_id)

#see subseting process. splitting off here bc the loading of epigen is slow. 
# all_da_results_subset_filter = find_interested_features(select_tissue_types)
# saveRDS(all_da_results_subset_filter, file = file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS"))
# all_da_results_subset_filter = readRDS(file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS")) 

all_da_results_subset_filter = readRDS(file.path(here(), "data", "glp1a_all_omes_subset_wz.RDS")) %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH"), .by = c("tissue", "assay", "comparison_group"))
```

Loading in data from the luminex assays and combining. The luminex assays had a much restricted number of features compared to the RNAseq, so I only pull GIP, GLP1, Glucagon.
```{r, eval = TRUE}
select_tissue_types_immuno = c(select_tissue_types, "PLASMA")
all_immuno_da_results = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types_immuno,
                                                                  assays = "IMMUNO",
                                                                  include_epigen = F) %>%
  filter(feature_ID %in% c("GIP", "GLP1", "GLUCAGON")) %>%
  mutate(gene_symbol = "") %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH"), .by = c("tissue", "assay", "comparison_group")) %>%
  select(feature_ID, gene_symbol, assay, tissue, sex, comparison_group, p_value, adj_p_value, logFC) %>%
  filter(adj_p_value < 0.1)

significant_gene_results = all_da_results_subset_filter %>%
  filter(adj_p_value < 0.1) %>%
  select(feature_ID, gene_symbol, assay, tissue, sex, comparison_group, p_value, adj_p_value, logFC) %>%
  arrange(assay)

full_significant_results = rbind(all_immuno_da_results, significant_gene_results)
# View(full_significant_results %>% filter(assay %in% c("TRNSCRPT", "IMMUNO")))
write.csv(full_significant_results, 
          file = file.path(here(), "data", "full_da_results.csv"),
          row.names = FALSE)
```

Protein check. None of the proteins have been sufficiently reliable detected for any of the incretin receptors, so we just avoid including proteomics in further analysis.
```{r}
all_da_results_subset = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types, assays = c("PROT")) %>%
  left_join(FEATURE_TO_GENE, by = "feature_ID") %>%
  filter(gene_symbol %in% interested_genes)
```
No longer used in the final manuscript. 
```{r figure 1 split by sex}
rna_tissues = select_tissue_types
immuno_tissues = MotrpacRatTraining6moData::TISSUE_ABBREV[!MotrpacRatTraining6moData::TISSUE_ABBREV %in% c("OVARY", "TESTES")]

# Glp1r & GLP1
rna_glp1r = get_expression_stats_by_sex("ENSRNOG00000001152", rna_tissues, "TRNSCRPT")

#we're just tossing plasma into blood to match up levels with RNA
immuno_glp1 = get_expression_stats_by_sex("GLP1", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

glp1_fig1 = plot_rna_immuno_jitterbysex(
  rna_data = rna_glp1r,
  immuno_data = immuno_glp1,
  include_first_legend = FALSE,
  gene_name = "GLP1R",
  rna_y_breaks = c(c(9,70)),
  immuno_reverse = FALSE
)

pdf(file = file.path(here(), "figures", "incretin_by_tissue", "fig_Supplemental_glp1_by_sex.pdf"),
    width = 10, height = 8)
print(glp1_fig1)
dev.off()

# Gipr & GIP
rna_gipr = get_expression_stats_by_sex("ENSRNOG00000015860", rna_tissues, "TRNSCRPT")
#we're just tossing plasma into blood to match up levels with RNA
immuno_gip = get_expression_stats_by_sex("GIP", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))

gip_fig1 = plot_rna_immuno_jitterbysex(
  rna_data = rna_gipr,
  immuno_data = immuno_gip,
  gene_name = "GIPR",
  immuno_reverse = FALSE
)

pdf(file = file.path(here(), "figures", "incretin_by_tissue", "fig_Supplemental_gip_by_sex.pdf"),
    width = 10, height = 8)
print(gip_fig1)
dev.off()
```

Accompanying glucagon expression check for manuscript.
```{r}
immuno_gcg = get_expression_stats_by_sex("GLUCAGON", immuno_tissues, "IMMUNO") %>%
  mutate(Tissue = case_when(
    Tissue == "PLASMA" ~ "BLOOD",
    TRUE ~ Tissue
  ))
```


```{r, include = FALSE}
other_genes = genes_in %>%
  dplyr::filter(!tolower(gene_name) %in% c("glp1r", "gipr")) %>%
  dplyr::distinct(gene_id, gene_name)

for (i in seq_len(nrow(other_genes))) {
  gene_id = other_genes$gene_id[i]
  gene_name_val = other_genes$gene_name[i]
  #not particularly efficient bc it loads everything for each call for the other genes... 
  rna_data = get_expression_stats_by_sex(gene_id, rna_tissues, "TRNSCRPT")
  if (nrow(rna_data) == 0) next

  gene_plot = plot_rna_immuno_jitterbysex(
    rna_data = rna_data,
    gene_name = gene_name_val,
    include_first_legend = TRUE
  )

  pdf(file = file.path(here(), "figures", "incretin_by_tissue", paste0("fig_Supplemental_", tolower(gene_name_val), "_by_sex.pdf")))
  print(gene_plot)
  dev.off()
}
```


```{r figure 2A, eval = T, warning = F}
scale = 2
top_selection = c("CORTEX", 
                  "HYPOTH",
                  "HIPPOC",
                  "BAT",
                  "WAT-SC",
                  "SKM-GN",
                  "SKM-VL"
)
bottom_selection = c("BLOOD", 
                     "HEART",
                     "LUNG",
                     "SPLEEN",
                     "ADRNL", 
                     "COLON", 
                     "SMLINT")

heatmap_trimmed = make_a_plot_for_tissue_heatmaps(top_selection,
                                                  interested_genes,
                                                  legend_tissue_types = c(top_selection, bottom_selection))
other_heatmap_trimmed = make_a_plot_for_tissue_heatmaps(bottom_selection,
                                                        interested_genes,
                                                        show_heatmap_legend = FALSE)
heatmap_full = heatmap_trimmed$ht %v% other_heatmap_trimmed$ht #concat vertically

#size the page to the drawn heatmap so there is no surrounding whitespace
zscore_draw_args = list(heatmap_full,
                        heatmap_legend_list = list(heatmap_trimmed$lgd),
                        merge_legends = FALSE,
                        align_heatmap_legend = "heatmap_top",
                        align_annotation_legend = "heatmap_top",
                        ht_gap = unit(12, "mm"))
zscore_dims = measure_heatmap(zscore_draw_args)

pdf(file = file.path(here(), "figures", "fig_2a_attempts", "all_tissue_combined_rna_zscore.pdf"),
    width = zscore_dims["width"], height = zscore_dims["height"], family = "ArialMT")
do.call(draw, zscore_draw_args)
dev.off()
```

```{r figure 2A logFC, eval = T, warning = F}
heatmap_trimmed_logfc = make_a_plot_for_tissue_heatmaps(top_selection,
                                                        interested_genes,
                                                        value_col = "logFC",
                                                        legend_tissue_types = c(top_selection, bottom_selection))
other_heatmap_trimmed_logfc = make_a_plot_for_tissue_heatmaps(bottom_selection,
                                                               interested_genes,
                                                               value_col = "logFC",
                                                               show_heatmap_legend = FALSE)
heatmap_full_logfc = heatmap_trimmed_logfc$ht %v% other_heatmap_trimmed_logfc$ht

logfc_draw_args = list(heatmap_full_logfc,
                       heatmap_legend_list = list(heatmap_trimmed_logfc$lgd),
                       merge_legends = FALSE,
                       align_heatmap_legend = "heatmap_top",
                       align_annotation_legend = "heatmap_top",
                       ht_gap = unit(12, "mm"))
logfc_dims = measure_heatmap(logfc_draw_args)

pdf(file = file.path(here(), "figures", "fig_2a_attempts", "all_tissue_combined_rna_logfc.pdf"),
    width = logfc_dims["width"], height = logfc_dims["height"], family = "ArialMT")
do.call(draw, logfc_draw_args)
dev.off()
```

```{r figure 2B triangle plots, eval = T}
cutoff = 0.10
#----here we can just find a big table of DA results if you'd like to view the tables.----
rna_selected_da_results = MotrpacRatTraining6mo::combine_da_results(tissues = select_tissue_types, assays = "TRNSCRPT") %>%
  filter(feature_ID %in% all_da_results_subset_filter$feature_ID) %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH"), .by = c("tissue", "assay", "comparison_group")) %>%
  filter(comparison_group == "8w") %>%
  select(assay, tissue, feature_ID, sex, comparison_group, adj_p_value, logFC) %>%
  pivot_wider(names_from = sex, values_from = c(adj_p_value, logFC), names_glue = "{sex}_{.value}") %>%
  mutate(shape = ifelse(male_logFC > 0 & female_logFC > 0, "up", "down")) %>%
  mutate(group = case_when(
    male_adj_p_value < cutoff & (is.na(female_adj_p_value) | female_adj_p_value >= cutoff) ~ "male-specific",
    female_adj_p_value < cutoff & (is.na(male_adj_p_value) | male_adj_p_value >= cutoff) ~ "female-specific",
    female_adj_p_value < cutoff & male_adj_p_value < cutoff  ~ "sex-consistent",
    TRUE ~ "neither-significant"
  ))  %>% 
  pivot_longer(cols = starts_with("male_") | starts_with("female_"), 
               names_to = c("sex", ".value"), 
               names_pattern = "(male|female)_(.*)")
#
rna_selected_da_results = merge(rna_selected_da_results, FEATURE_TO_GENE) %>%
  mutate(group = factor(group,
                        levels = c("sex-consistent", "female-specific", "male-specific", "neither-significant")))

#Because of the shapes the order cant really be forced to be in the same order as the other heatmaps?. 
#Going to stop troubleshooting because it is taking way too long.
rna_selected_da_results$gene_symbol = factor(rna_selected_da_results$gene_symbol,
                                             levels = sort(unique(rna_selected_da_results$gene_symbol)))


triangle_plots = ggplot(
  rna_selected_da_results,
  aes(y = gene_symbol, x = tissue, size = group, shape = shape)
) +
  geom_point(aes(fill = group, colour = group)) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(),
    plot.background = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(linewidth = 1),
    axis.text.y = element_text(colour = "black"),
    axis.text.x = element_text(
      colour = "black",
      angle = 90, hjust = 1, vjust = 0.5
    ),
    axis.title = element_blank(),
    strip.text.y.left = element_text(size = 10),
    strip.text.x = element_text(angle = 90, hjust = 0),
    axis.text.y.right = element_text(hjust = 0),
    legend.position = "top",
    legend.box = "vertical",
    legend.justification = "right",
    legend.title = element_blank(),
    legend.margin = margin(t = 0, b = -0.3, unit = "cm"),
    legend.spacing.x = unit(0.1, "cm")
  ) +
  scale_fill_manual(
    values = c(
      "sex-consistent" = "white",
      "male-specific" = MotrpacBicQC::sex_cols[["male"]],
      "female-specific" = MotrpacBicQC::sex_cols[["female"]]
    ),
    breaks = c("sex-consistent", "female-specific", "male-specific"),
    name = "Node type"
  ) +
  scale_colour_manual(
    values = c(
      "sex-consistent" = "black",
      "male-specific" = MotrpacBicQC::sex_cols[["male"]],
      "female-specific" = MotrpacBicQC::sex_cols[["female"]]
    ),
    breaks = c("sex-consistent", "female-specific", "male-specific"),
    name = "Node type"
  ) +
  scale_shape_manual(
    values = c(up = 24, down = 25),
    labels = c(up = "Up-regulated", down = "Down-regulated"),
    name = "Node direction"
  ) +
  scale_size_manual(
    values = c(
      "sex-consistent" = 4,
      "female-specific" = 3,
      "male-specific" = 2.5
    ),
    breaks = c("sex-consistent", "female-specific", "male-specific"),
    name = "Node type"
  ) +
  guides(
    shape = guide_legend(
      override.aes = list(size = 3, fill = "black")
    ),
    size = guide_legend(
      override.aes = list(
        shape = 24,
        size = c(3, 2.5, 1.5),
        fill = c(
          "white",
          MotrpacBicQC::sex_cols[["female"]],
          MotrpacBicQC::sex_cols[["male"]]
        ),
        colour = c("black", "white", "white")
      )
    )
  ) +
  ggtitle("Responses after 8 weeks of training")


pdf(file = paste0(here(), "/figures/fig_2c_rna_triangle_direction.pdf"),
    width = 6, height = 4)
print(triangle_plots)
dev.off()
```

Making the immuno heatmaps but instead we're just doing timepoint by tissue
```{r figure 3 Immuno heatmaps, message = F}
immuno_tissue_types = c(select_tissue_types, "PLASMA")

shared_color_range = c(-3, 3)

# GIP carries the legend; GLP1 suppresses its duplicate
gip_heatmap = make_immuno_heatmap("GIP", color_range = shared_color_range, show_legend = TRUE, show_anno_names = FALSE)
glp_heatmap = make_immuno_heatmap("GLP1", color_range = shared_color_range, show_legend = FALSE, show_anno_names = TRUE)

combined_heatmap = gip_heatmap$plot + glp_heatmap$plot

immuno_draw_args = list(combined_heatmap, annotation_legend_list = gip_heatmap$legend)
immuno_dims = measure_heatmap(immuno_draw_args)

pdf(file = paste0(here(), "/figures/fig_3_gip_glp_immunoassay.pdf"),
    width = immuno_dims["width"], height = immuno_dims["height"])
do.call(draw, immuno_draw_args)
dev.off()


all_immuno_da_results = MotrpacRatTraining6mo::combine_da_results(tissues = immuno_tissue_types,
                                                                    assays = "IMMUNO",
                                                                    include_epigen = F) %>%
    filter(feature_ID %in% c("GLP1")) %>%
    mutate(adj_p_value = p.adjust(p_value, method = "BH"), .by = c("tissue", "assay", "comparison_group")) %>%
    select(tissue, sex, comparison_group, logFC) %>%
    tidyr::pivot_wider(id_cols = tissue,
                       names_from = c(sex, comparison_group),
                       values_from = logFC) %>%
    column_to_rownames("tissue")
```

Figure 1E for revisions

GTEx human reference expression for the incretin receptor genes. BAT and VENACV have no GTEx equivalent and are excluded. SKM-GN and SKM-VL both map to Muscle_Skeletal.
```{r figure 1E expression presence table}
rat_for_poly = get_motrpac_expression(interested_genes, 
                                      rna_tissues,
                                      motrpac_da = all_da_results_subset_filter)
rat_to_gtex = c(
  "ADRNL" = "Adrenal_Gland",
  "BLOOD" = "Whole_Blood",
  "COLON" = "Colon_Sigmoid",
  "CORTEX" = "Brain_Cortex",
  "HEART" = "Heart_Left_Ventricle",
  "HIPPOC" = "Brain_Hippocampus",
  "HYPOTH" = "Brain_Hypothalamus",
  "KIDNEY" = "Kidney_Cortex",
  "LIVER" = "Liver",
  "LUNG" = "Lung",
  "SKM-GN" = "Muscle_Skeletal",
  "SKM-VL" = "Muscle_Skeletal",
  "SMLINT" = "Small_Intestine_Terminal_Ileum",
  "SPLEEN" = "Spleen",
  "WAT-SC" = "Adipose_Subcutaneous"
)
#prefer to have more suitable tissue groupings
tissue_order = c(
  "CORTEX",
  "HYPOTH", 
  "HIPPOC", 
  "BAT",
  "WAT-SC",
  "SKM-GN",
  "SKM-VL",
  "BLOOD", 
  "HEART",
  "VENACV",
  "LUNG", 
  "SPLEEN",
  "KIDNEY",
  "LIVER",
  "ADRNL",
  "COLON",
  "SMLINT"
)

gtex_for_poly = get_gtex_expression(interested_genes, rat_to_gtex, select_tissue_types)

poly_data = dplyr::bind_rows(
  make_triangle_poly(rat_for_poly, tissue_order, interested_genes),
  make_triangle_poly(gtex_for_poly, tissue_order, interested_genes)
)

disagree_cells = dplyr::inner_join(
  rat_for_poly %>% dplyr::select(tissue, gene, rat_detected = detected),
  gtex_for_poly %>% dplyr::select(tissue, gene, gtex_detected = detected),
  by = c("tissue", "gene")
) %>%
  dplyr::filter(!is.na(rat_detected) & !is.na(gtex_detected) & rat_detected != gtex_detected) %>%
  dplyr::mutate(
    x_num = match(tissue, tissue_order),
    y_num = match(gene, interested_genes)
  ) %>%
  dplyr::filter(!is.na(x_num))

combined_split_plot = ggplot(poly_data, aes(x = px, y = py, group = group, fill = fill_val)) +
  geom_polygon() +
  geom_rect(
    data = disagree_cells,
    aes(xmin = x_num - 0.5, xmax = x_num + 0.5, ymin = y_num - 0.5, ymax = y_num + 0.5,
        color = "Species disagree"),
    fill = NA,
    linewidth = 1,
    inherit.aes = FALSE
  ) +
  geom_hline(yintercept = seq(0.5, length(interested_genes) + 0.5, 1),
             color = "grey70", linewidth = 0.3) +
  geom_vline(xintercept = seq(0.5, length(tissue_order) + 0.5, 1),
             color = "grey70", linewidth = 0.3) +
  scale_x_continuous(
    breaks = seq_along(tissue_order),
    labels = tissue_order,
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq_along(interested_genes),
    labels = interested_genes,
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("detected" = "#d4edda", "not_detected" = "#f8d7da", "no_data" = "grey85"),
    labels = c("detected" = "Detected", "not_detected" = "Not detected", "no_data" = "No GTEx data"),
    name = NULL
  ) +
  scale_color_manual(
    values = c("Species disagree" = "black"),
    name = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 10, color = "black"),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.position = "top"
  ) +
  ggtitle("Incretin Expression in Rat (upper right) and Human (lower left)")

pdf(file = file.path(here(), "figures","final_figs", "Fig_1C_combined_split_detection_table.pdf"),
    width = 16, height = 5)
print(combined_split_plot)
dev.off()
```

```{r}
cat(summarize_detection(rat_for_poly, gtex_for_poly, tissues = tissue_order))
```

```{r}
detection_summary = compute_detection_summary(rat_for_poly, gtex_for_poly, tissues = tissue_order) %>%
  mutate(
    rat = paste0(rat_n_detected, "/", n_genes, " (", rat_pct_detected, "%)"),
    human = paste0(human_n_detected, "/", n_comparable, " (", human_pct_detected, "%)"),
    agreement = paste0(n_agree, "/", n_comparable, " (", pct_agree, "%)")
  ) %>%
  select(tissue, rat, human, agreement)
n_rows = nrow(detection_summary)
tbl = detection_summary %>%
  kable(
    col.names = c("Tissue", "Rat detected", "Human detected", "Agreement"),
    caption = "<span style=\"color:black;font-size:24px;\">Incretin receptor detection and rat-human agreement by tissue</span>",
    escape = FALSE
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, font_size = 14) %>%
  row_spec(0, bold = TRUE, font_size = 18, color = "black") %>%
  row_spec(1:n_rows, color = "black") %>%
  column_spec(1:4, extra_css = "padding: 6px 14px; color: black;")

tbl

fig_1b_path = here("figures/final_figs/Fig_1B_detection_summary_table.png")
save_kable(tbl, file = fig_1b_path, zoom = 2, density = 900)

#pad the saved table with an outer white margin (taller on top) so the composite
#panel label (B) sits above the caption and never overlaps the table text
fig_1b_img = magick::image_read(fig_1b_path) %>%
  magick::image_background("white") %>%
  magick::image_border("white", "60x170")
magick::image_write(fig_1b_img, fig_1b_path)
```






